\chapter*{Preface}
\pagestyle{fancy}%

\firstchar{T}his book explains the basics of text processing in Java.  
Java provides an extensive set of tools for manipulating
character-based data, including byte-to-character conversion, input
and output streams including sockets and compression, memory-mapping
files, and high-level methods for manipulating sequences of characters
like buffers and regular expressions.

This book assumes you are familiar with basic coding in a language like
Java, but that you may not be familiar with low-level representations
of bytes and integers, the representation and manipulation of
strings, how to access objects from the classpath as resources,
or how to write non-greedy regular expressions based on Unicode.

The presentation here will be hands on.  You should be comfortable
reading short and relatively simple Java programs.  Java programming
idioms like loop boundaries being inclusive/exclusive and higher-level
design patterns like visitors will also be presupposed.  More specific
aspects of Java coding relating to text processing, such as streaming
I/O, character decoding, string representations, and regular
expression processing will be discussed in more depth.  We will also
go into some detail on collections, XML/HTML parsing with SAX, and
serialization patterns.  Open-source software dependencies include the
International Components for Unicode (ICU) library and Apache Lucene
search library.  We provide a warmup chapter on using the various
tools required for text editors to version control to the Ant build
system for Java.  As a finale, we cover the use of the Apache Lucene
search library and the Apache Solr search server.

This book grew out of work on its companion volume, {\it Natural
Language Processing with LingPipe}.  We found that the background material
on Java was growing to the point where it made more sense to pull it
out into its own standalone package.  We have included a few examples
of basic text processing using LingPipe. 

We hope you have as much fun with text processing in Java as we've
had.  We'd love to hear from you, so feel free to drop us a line at
\urlin{info@lingpipe.com} with any kind of question or comment you
might have.

\vspace*{0.125in}

\newlength{\sigWidth}
\settowidth{\sigWidth}{{\small\sc September 30, 2525}}
\mbox{ }
\hfill
\parbox{\sigWidth}{
Bob Carpenter
\\[2pt]
Mitzi Morris
\\[2pt]
Breck Baldwin
\\[2pt]\small
\\
\it New York
\\
\today}

